Propositions as Sessions, Mechanically

### 2015-04-09 paper/formalisation.tex
Mostly finished CP section (good enough for second draft).

Figure~\ref{fig:cll} presents the well-typed process relation for CP as an
inductive type in Coq. Each typing rule in CP is represented as a constructor
of the inductive type \coqe$cp_rule$. The inductive type is indexed by a
process (\coqe$proc$) and an environment (\coqe$penv$) indicated by them
appearing after the colon on the first line. \coqe$forall$ quantifies over
hypotheses in a constructor; everything before the comma. Compound hypotheses
are given labels, such as \key{PER}, which is the default name used to
introduce them into the proof context during theorem proving. Some details
such as environment uniqueness have been elided, but notice the use of LNCQ
form within the hypotheses of the rules with binders. These rules exclude a
finite set of names \coqe$L$ from opening the subprocesses where opening a
process \coqe$P$ with name \coqe$x$ is denoted by \coqe$open_proc P x$. Bound
variables are represented using de Bruijn indices and free variables are
represented using an abstract type, \coqe$atom$, provided by the Metatheory
library. The rules with binders specify the type of a bound variable in the
constructor rather than a name. Note that the rules for polymorphism,
$\exists$ and $\forall$, were not included in the final Coq development since
GV (\S~\ref{sec:gv}) does not support polymorphism making the rules
unnecessary for the translation. Further, as noted by \citeauthor{Lee:2012},
support for polymorphism in type variables creates an explosion of the
infrastructure requiring additional of opening and substituting definitions
and lemmas.

@@

where $\open{k}{x}{P}$ denotes replacing de Bruijn index $k$ with free
variable $x$ in process $P$.

@@

% ...commuting conversions

%\input{cpstructeq}

%\input{commconvonecp}

%\input{commconvtwocp}

The structural equivalences and commuting conversions are shown in
Figure~\ref{??} and Figures~\ref{??} and~\ref{??}, respectively. These rules
are expressed to indicate some additional complexity not present in the
original presentation. For example, structural equivalences are not unique and
different process terms result depending on initial binder
permutation. $\permb{0}{1}$ indicates the permuting of binders $0$ and $1$ in
process $\tm{P}$. The paper presentation specifies only one possible
associativity rule lending to the specific split of the linear environments.
Notice that permutating binders is essential to formulating the majority of
these rules. Analysing permutations containing singleton bindings was
necessary for the commuting conversion rules. Singleton bindings can occur
frequently during a proof and it is sometimes necessary to analyse a
permutation of the form P++x~A ~~ Q++R to determine whether x occurs in Q or
R. the rule for associativity requires this and so do quite a few reduction
rules. since this involves some element of searching the context followed by
destructing hypotheses it could be effectively automated in permutations in
subprocesses which use a binder of the parent. the rule for associativity w
linear environments caused a problem with the splitting up of variables.

@@

The theorem for top-level cut elimination can be expressed as:
\begin{coq}
Theorem proc_cut_elim:
  forall P $\Gamma$
         (WT: P $\tpcp$ $\Gamma$),
    exists Q, P $\becomes^\star$ Q /\ $\sim$ is_cut Q.
\end{coq}

where $\becomes^\star$ denotes the reflexive and transitive closure of
$\becomes$ and $\sim$ \coqe$is_cut Q$ expresses that Q is not a cut.

>>>>

Figure~\ref{fig:cll} presents the well-typed process relation for CP as an
inductive type in Coq. Each typing rule in CP is represented as a constructor
of the inductive type \coqe$cp_rule$. The inductive type is indexed by a
process (\coqe$proc$) and an environment (\coqe$penv$) indicated by them
appearing after the colon on the first line. \coqe$forall$ quantifies over
hypotheses in a constructor; everything before the comma. Compound hypotheses
are given labels, such as \key{PER}, which is the default name used to
introduce them into the proof context during theorem proving. Some details
such as environment uniqueness have been elided, but notice the use of LNCQ
form within the hypotheses of the rules with binders. These rules exclude a
finite set of names \coqe$L$ from opening the subprocesses where opening a
process \coqe$P$ with name \coqe$x$ is denoted by \coqe$open_proc P x$ and
corresponds to replacing de Bruijn index 0 with \coqe!x!. Bound variables are
represented using de Bruijn indices and free variables are represented using
an abstract type, \coqe$atom$, provided by the Metatheory library. The rules
with binders specify the type of a bound variable in the constructor rather
than a name. Note that the rules for polymorphism, $\exists$ and $\forall$,
were not included in the final Coq development since GV (\S~\ref{sec:gv}) does
not support polymorphism making the rules unnecessary for the
translation. Further, as noted by \citeauthor{Lee:2012}, support for
polymorphism in type variables creates an explosion of the infrastructure
requiring additional of opening and substituting definitions and lemmas.

@@

where $\open{k}{x}{P}$ denotes replacing de Bruijn index $k$ with free
variable $x$ in process $P$~\footnote{For example, \coqe!$\open{0}{x}{P}
  \defeq$ open_proc P x!}.

@@

The structural equivalences and commuting conversions have some additional
complexity not present in the original presentation. For instance, structural
equivalences are not unique and different process terms result depending on
the initial binder permutation. Consider the associativity rule for which the
paper presentation specifies only one rule:
\begin{coq}
Lemma assoc_1:
  forall P Q R A B $\Gamma$
         (LCP: forall x, lc_proc (open_proc P x))
         (WT: $\nu$ B.($\nu$ A.(P $\mid$ Q) $\mid$ R) $\tpcp$ $\Gamma$),
    $\nu$ A.(P $\mid$ $\nu$ B.($\permb{0}{1}{Q}$ $\mid$ R)) $\tpcp$ $\Gamma$.
\end{coq}

where \coqe$lc_proc P$ specifies that \coqe$P$ is a \textit{local closure}, in
other words contains no unbound de Bruijn indices and $\permb{0}{1}Q$
indicates the permuting of binders 0 and 1 in process $\tm{Q}$. The typing
rule for \coqe$cp_cut$ does not specify any particular split of the linear
environment in the conclusion; splitting is non-deterministic. So, in order to
prove the above statement, it must be true in \key{WT} that \coqe$Q$ takes the
binder of type \coqe$B$ and not \coqe$P$, hence the local closure assumption
on \coqe$open_proc P x$. Similarly, to prove the converse statement
\coqe$forall x, open_proc R x$ must be assumed. For commuting conversions,
permutating binders is essential in formulating the majority of the rules.

Regarding commuting conversions, many require analysing permutations
containing singleton bindings. Consider the following commuting conversion:
\begin{coq}
  Lemma reduce_cc_input:
    forall P Q (x:atom) A B $\Gamma$
           (WT: $\nu$ A.(x(B).P $\mid$ Q) $\tpcp$ $\Gamma$),
      x(B).($\nu$ A.($\open{0}{1}{P}$ $\mid$ Q)) $\tpcp$ $\Gamma$.
\end{coq}

Afer inverting \key{WT} and specialising the resulting hypotheses, regarding
the subprocesses, with fresh variables the proof state is as follows:

\begin{tabular}{l}
\begin{coq}
1 subgoal
(* ... other elements of the proof context ... *)
CPQ : open_proc Q y $\tpcp$ y$\sim$A++$\Delta$Q
CPP0 : $\open{0}{z}{\open{1}{y}{P}}$ $\tpcp$ z$\sim$B++x$\sim$C++$\Delta$P0
PER : Permutation $\Gamma$ ($\Delta$P++$\Delta$Q)
PER0 : Permutation ($\Delta$P++y$\sim$A)]) (x$\sim$B$\parr$C++$\Delta$P0)
\end{coq}\\ \hline
\begin{coq}
x(B).($\nu$ A.($\open{0}{1}{P}$ $\mid$ Q)) $\tpcp$ $\Gamma$.
\end{coq}
\end{tabular}
\\~\\

In order to prove the goal, $\Gamma$ must be replaced with the sublists typing
\coqe!P! and \coqe!Q!. Therefore, \coqe!PER0! needs to modified such that the
left-hand side is $\Delta$P and the right-hand side has \coqe!y$\sim$A!
removed. In other words, the binding of \coqe!y! within \coqe!$\Delta$P0! must
be extracted. Since many of the commuting conversion rules required this sort
of reasoning separating the work out into a tactic comes to mind. Indeed, Ltac
is well-suited for this type of work, searching the context and manipulating
hypotheses of a particular form. The Ltac script in Figure~\ref{fig:extract}
does the job\todo{keep (and explain) this?}.

\begin{figure}
\begin{coq}
Ltac extract_bnd x A :=
  match goal with
    | [H: Permutation (_++x~A) ?F |- _] =>
      let BNDS := fresh "BNDS" in
      forwards~ BNDS: Perm_binds x A H
      ; analyze_binds_uniq BNDS
      ; try (by applys~ uniq_perm H)
      ; try (match goal with
               | [H: binds x A ?E |- _] =>
                 let E1 := fresh "E1" in
                 let E2 := fresh "E2" in
                 let EQ := fresh "EQ" in
                 apply binds_env_split in H
                 ; inversion_clear H as (E1 & E2 & EQ)
                 ; substs~; des_reqs
             end)
  end.
\end{coq}
\caption{Tactic finds binding in permutation RHS}
\label{fig:extract}
\end{figure}

However, the task of removing the binding from the permutation is not
automated to keep information loss within the context transparent to an
individual proof. The tactic appears more general in this form and allows more
than one ordering of tactic sequences. feels like two jobs to remove it as
well as find it. although perhaps extract is not the right term if it doesn't
remove it

@@

\begin{figure}
\begin{coq}
Theorem proc_cut_elim:
  forall P $\Gamma$
         (WT: P $\tpcp$ $\Gamma$),
    exists Q, P $\becomes^\star$ Q /\ $\sim$ is_cut Q.
\end{coq}
\caption{Cut elimination theorem}
\label{fig:cutelim}
\end{figure}

The theorem for top-level cut elimination appears in Figure~\ref{fig:cutelim}
where $\becomes^\star$ denotes the reflexive and transitive closure of
$\becomes$ and $\sim$ \coqe$is_cut Q$ expresses that Q is not a cut.
